Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2022-01-26 fstats_pca_lewontin.tex

\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics \textit{sensu} Patterson. I show that these two perspectives are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that the first few PCs are sufficient to approximate most $F$-statistics, and thus PCA-plots are effective to predict $F$-statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete descriptions of human genetic variation. 
\end{abstract}
\section{Introduction}
As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2017, witt2022}. In turn, an important goal of population genetics is to use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.

@@

\paragraph{F-statistics}
When the number of populations exceeds a few dozen, even codifying reasonable population models can be prohibitively difficult. One approach is to pick a small set of ``representative'' samples, and restrict modeling to this subset \citep[e.g.][]{gravel2011, harney2021}. However, this has the drawback that a large proportion of the available data remains unused. An increasingly popular alternative approach, particularly in the analysis of human ancient DNA, is therefore to focus on the relationship between two, three or four populations, commonly using $F$-statistics \textit{sensu} Patterson \citep{reich2009, patterson2012, peter2016}. Formal definition will be given in the Theory section; but an informal motivation starts with the null model that populations are related as a tree, in which each $F$-statistic measures the length of a particular set of branches. \citep[Figure \ref{fig:geom};][]{semple2003, peter2016}.

@@

The most widely considered alternative model is an admixture graph \citep{patterson2012}, an example is given in Figure \ref{fig:admix}A. Here, the *typically unobserved) population $X_y$ is generated by a mixture of individuals from the ancestors of $X_2$ and $X_3$. Over time, genetic drift will change $X_y$ to $X_x$, which is the admixed population we observe. In this case, all $F_4$-statistics involving $X_y$ and both admixture sources will be non-zero, and, in some cases, $F_3(X_y; X_2, X_3)$ will be negative \citep[exact conditions can be found in ][]{peter2016}.

\subsubsection{Geometric interpretation of $F$-statistics}
An implicit assumption in the development of $F$-statistics is that population lineages are mostly discrete, and that gene flow is rare. Specifically, they interpret the populations $X_i$ as points or vectors in the $S$-dimensional \emph{allele frequency space} $\mathbb{R}^S$. In this case, the $F$-statistics can be thought of as inner (or dot) products, and they showed that all properties and tests related to treeness can be derived in this larger space. In particular the $F$-statistics can be written as
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
	\begin{align}
	\langle a, b \rangle &= \sum_i a_ib_i\\
	\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
	\langle a, a \rangle &= \normsq{a}\\
	\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
	\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$. The inner product is  closely related to vector projections
\begin{equation}
\vectorproj[b]{a} = \frac{\langle a , b\rangle}{\normsq{b}} b,\label{eq:proj}
\end{equation}
which is a vector colinear to $b$ whose length measures how much vector $a$ points in the direction of $b$. Thinking of $F$-statistics as projections also holds on trees: In e.g. a $F_4(X_1, X_4; X_2, X_3)$-statistic (Figure \ref{fig:admix}C), the internal branch is precisely the intersection of the paths from $X_1$ to $X_4$ and from $X_2$ and $X_3$. The external branches are independent populations, and so they are expected to drift orthogonally to each other.

@@

Let us again assume we have allele frequency data as above, but let us now assume we aggregate the allele frequency vectors $X_i$ of many populations in a matrix $\MX$ whose entry $x_{il}$ reflects the allele frequency of the $i$-th population at the $l$-th genotype. If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$. Since the allele frequencies are between zero and one, we can interpret each Population $X_i$
of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}, which is a subset of $\mathbb{R}^S$.

One way PCA can be motivated is that it aims to find a $K$-dimensional subspace of the data space that retains most  variation in the data. $K$ is at most $n-1$, in which case the data is simply rotated. However, the  historical processes that generated genetic variation often result in \emph{low-rank} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.

@@

\begin{figure}[!ht]
	\includegraphics[width=.6\textwidth]{figures/outgroup_f3.pdf}
	\caption{\textbf{Outgroup-$F_3$-statistics}. Interpretation of outgroup $F_3$-statistic on a tree (A) and PCA-plot (B).  The red segment represents $F_3(X_O; X_U, X_3)$ and the dashed blue segment reflects $F_3(X_O; X_U, X_1)$ and $F_3(X_O; X_U, X_2)$, which have the same value. 
	}
	\label{fig:outgroupf3}
\end{figure}
\subsection{Outgroup-$F_3$-statistics as projections}
A common application of $F_3$-statistics is, given an unknown sample $X_U$, to find the most closely related population among a reference panel $(X_i)$ \citep{raghavan2014}. This is done  using an \emph{outgroup}-$F_3$-statistic $F_3(X_O; X_U, X_i)$, where $X_O$ is an outgroup. The reason an outgroup is introduced is to account for differences in sample times and additional drift in the reference populations (\ref{fig:outgroupf3}A) The outgroup-$F_3$-statistic $F_3(X_O; X_U, X_3)$ represents the branch length from $X_O$ to the common node between the three samples in the statistic, and the closer this node is to $X_U$, the longer the branch and hence the larger the $F_3$-statistic.

@@

The two data sets I analyzed here suggest that two PCS (for the World data set) respectively three PCS (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:f3}E, \ref{fig:f4}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure. On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:f3}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs. Thus, I would also expect individual-based analyses and $F_4$-statistics between closely related populations to require substantially more PCs.

@@

\subsubsection{Missing data}
The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing have been proposed \citep[e.g.][]{hastie2015, meisner2018, meisner2021}. Calculating $F$-statistics first is yet another way to impute data for PCA that works well if missingness is low \citep{meisner2021}. In contrast, missing data in $F$-statistics is handled by estimating a standard error using resampling across the genome \citep{patterson2012}, which does not distinguish between biological and sampling variation. These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}, and a PCA-based approach to estimating $F$-statistics may provide higher accuracy than naive approaches because it allows using data from related populations.

\subsubsection{Normalization}
The normalization of SNPs is similarly a matter of convention. The $F$-statistic framework assumes that each SNP is an identically-distributed (but not independent) random variable; and the same would hold if SNPs were weighted. The drawback for individual $F$-statistics is that this adds a dependency on additional samples (through the mean allele frequency) that may be unwanted for individual $F$-statistics, but could be advantageous for tools that aim to do joint inference from  many $F$-statistics \citep{patterson2012, harney2021}. As genetic differentiation between human populations is low, the  normalization used may matter little in practice, but should be explored in future work \citep{felsenstein1973}.

@@

This continuum between admixture and population structure becomes more problematic when interpreting tools that  use multiple $F$-statistics to build compound models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}. One issue with these approaches is that they are usually restricted to at most a few dozen populations. As ancient DNA data sets now commonly include thousands of individuals, analysts are faced with the challenge of which data to include. A common approach is to sample a large number of distinct models, and retain the ones that are compatible with the data. However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits. The PCA-based  interpretation of $F$-statistics offers an alternative that trades interpretability for robustness, as I do not to assume that gene flow is rare, and visualizing both the projection used for estimating admixture proportion, and the residual space, can be used to highlight scenarios where the assumptions are violated.

@@

\newpage
%\bibliography{main}
\printbibliography

>>>>

\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics \textit{sensu} Patterson. I show that these two perspectives are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that the first few PCs are sufficient to approximate most $F$-statistics, and thus PCA-plots are effective to predict $F$-statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete descriptions of human genetic variation. 
\section{Introduction}
As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2017, witt2022}. In turn, an important goal of population genetics is to use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.

@@

When the number of populations exceeds a few dozen, even codifying reasonable population models can be prohibitively difficult. One approach is to pick a small set of ``representative'' samples, and restrict modeling to this subset \citep[e.g.][]{gravel2011, harney2021}. However, this has the drawback that a large proportion of the available data remains unused. An increasingly popular alternative approach, particularly in the analysis of human ancient DNA, is therefore to focus on the relationship between two, three or four populations, commonly using $F$-statistics \textit{sensu} Patterson \citep{reich2009, patterson2012, peter2016}. Formal definition will be given in the Theory section; but an informal motivation starts with the null model that populations are related as a tree, in which each $F$-statistic measures the length of a particular set of branches. \citep[Figure \ref{fig:geom};][]{semple2003, peter2016}.

@@

The most widely considered alternative model is an admixture graph \citep{patterson2012}, an example is given in Figure \ref{fig:admix}A. Here, the (typically unobserved) population $X_y$ is generated by a mixture of individuals from the ancestors of $X_2$ and $X_3$. Over time, genetic drift will change $X_y$ to $X_x$, which is the admixed population we observe. In this case, all $F_4$-statistics involving $X_y$ and both admixture sources will be non-zero, and, in some cases, $F_3(X_y; X_2, X_3)$ will be negative \citep[exact conditions can be found in ][]{peter2016}.

\subsubsection{Geometric interpretation of $F$-statistics}
An implicit assumption in the development of $F$-statistics is that population lineages are mostly discrete, and that gene flow is rare. Specifically, they interpret the populations $X_i$ as points or vectors in the $S$-dimensional \emph{allele frequency space} $\mathbb{R}^S$. In this case, the $F$-statistics can be thought of as inner (or dot) products, and they showed that all properties and tests related to treeness can be derived in this larger space. In particular the $F$-statistics can be written as
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
	\begin{align}
	\langle a, b \rangle &= \sum_i a_ib_i\\
	\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
	\langle a, a \rangle &= \normsq{a}\\
	\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
	\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$. The inner product is  closely related to the vector projection
\begin{equation}
\vectorproj[b]{a} = \frac{\langle a , b\rangle}{\normsq{b}} b,\label{eq:proj}
\end{equation}
which is a vector colinear to $b$ whose length measures how much vector $a$ points in the direction of $b$. Thinking of $F$-statistics as projections also holds on trees: In e.g. a $F_4(X_1, X_4; X_2, X_3)$-statistic (Figure \ref{fig:geom}C), the internal branch is precisely the intersection of the paths from $X_1$ to $X_4$ and from $X_2$ and $X_3$. On trees, all disjoint paths are independent (i.e. orthogonal) from each other, and thus the external branches vanish under the projection.

@@

Let us again assume we have allele frequency data as above, but let us now assume we aggregate the allele frequency vectors $X_i$ of many populations in a matrix $\MX$ whose entry $x_{il}$ reflects the allele frequency of the $i$-th population at the $l$-th genotype. If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$. Since the allele frequencies are between zero and one, we can interpret each population $X_i$
of $\MX$ as a point in $\mathbb{R}^S$.

PCA allows us to approximate the points in the high-dimensional allele frequency space by a $K$-dimensional subspace of the  data. If all PCs are considered, $K=n-1$, in which case the data is simply rotated. However, the  historical processes that generated genetic variation often result in \emph{low-rank} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.

@@

\begin{figure}[!ht]
	\includegraphics[width=.6\textwidth]{figures/outgroup_f3.pdf}
	\caption{\textbf{Outgroup-$F_3$-statistics}. Interpretation of outgroup $F_3$-statistic on a tree (A) and PCA-plot (B).  The red segment represents $F_3(X_O; X_U, X_3)$ and the dashed blue segment reflects $F_3(X_O; X_U, X_1)$ and $F_3(X_O; X_U, X_2)$, which have the same value. 
	}
	\label{fig:outgroupf3}
\end{figure}
\subsection{Outgroup-$F_3$-statistics as projections}
A common application of $F_3$-statistics is, given an unknown sample $X_U$, to find the most closely related population among a reference panel $(X_i)$ \citep{raghavan2014}. This is done  using an \emph{outgroup}-$F_3$-statistic $F_3(X_O; X_U, X_i)$, where $X_O$ is an outgroup. The reason an outgroup is introduced is to account for differences in sample times and additional drift in the reference populations (Figure \ref{fig:outgroupf3}A). The outgroup-$F_3$-statistic $F_3(X_O; X_U, X_3)$ represents the branch length from $X_O$ to the common node between the three samples in the statistic, and the closer this node is to $X_U$, the longer the branch and hence the larger the $F_3$-statistic.

@@

The two data sets I analyzed here suggest that two PCs (for the World data set) respectively three PCs (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:f3}E, \ref{fig:f4}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure. On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:f3}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs. Thus, I would also expect individual-based analyses and $F_4$-statistics between closely related populations to require substantially more PCs.

@@

\subsubsection{Missing data}
The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing missing data have been proposed \citep[e.g.][]{hastie2015, meisner2018, meisner2021}. Calculating $F$-statistics first is yet another way to impute data for PCA that works well if missingness is low \citep{meisner2021}. In contrast, missing data in $F$-statistics is handled by estimating a standard error using resampling across the genome \citep{patterson2012}, which does not distinguish between biological and sampling variation. These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}. The theory developed here suggests that developing a PCA-based $F$-statistics framework might be more powerful, as related populations could be used to fill in missing data.

\subsubsection{Normalization}
In PCA, SNPs are typically normalized to have expected variance of one; a step that is omitted in calculating $F$-statistics \citep{patterson2006}. The $F$-statistic framework assumes that each SNP is an identically-distributed (but not independent) random variable. which holds regardless of weighting. Thus, normalization of SNPs is largely a matter of convention, for  $F$-statistics the dependency on  additional samples (through  mean allele frequencies) is often unwanted, but could be advantageous for tools that aim to do joint inference from  many $F$-statistics such as \texttt{qpAdm} \citep{patterson2012, harney2021}. As genetic differentiation between human populations is low, the  normalization used may matter little in practice, but should be explored in future work \citep{felsenstein1973}.

@@

This continuum between admixture and population structure becomes more problematic when interpreting tools that  use multiple $F$-statistics to build compound models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}. One issue with these approaches is that they are usually restricted to at most a few dozen populations. As ancient DNA data sets now commonly include thousands of individuals, analysts are faced with the challenge of which data to include. A common approach is to sample a large number of distinct models, and retain the ones that are compatible with the data. However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits. The PCA-based  interpretation of $F$-statistics offers an alternative that trades interpretability for robustness, as I do not need to assume that gene flow is rare.

@@

%\bibliography{main}
\printbibliography
